Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control

### 2023-08-18 Rmarkdown/paper.Rmd
minor edits to manuscript text

<!-- to do: -->
<!-- 1. cite https://doi.org/10.1080/00031305.2018.1549100 for bayesian r2? -->

@@

Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified. The best sentinel homes, then, are often short-lived. Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with median pH and dissolved inorganic carbon concentrations in pipe rack effluent of 7.3--7.4 and 4--5 mg C L^-1^ (Table 1). But while water quality in Zones 1 and 2 was largely similar, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022]. Aluminum concentrations in Zone 1 were much lower and more consistent throughout the year (Table S1).

@@

Doubling the orthophosphate dose decreased lead release from pipe racks substantially. At 1 mg PO~4~ L^-1^, lead in pipe rack effluent did not exhibit a clear time-trend  (Figure 1c). In particular, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ everywhere (d[Pb]/dt ~ 0, Figure 1a). Pipe racks, then, appear to have stabilized well at the initial orthophosphate concentration.

@@

Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding Zone 2-specific smooth term (Figure 1a). One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. While this did not impact distributed water pH, it did lower aluminum in treated water (Figure S6). And a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles [@kvech_role_2001]. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999]. But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data are not rich enough to identify the true cause.

@@

Temperature variation in the pipe rack effluent was a reasonable match for distributed water. While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked distributed water and point-of-use flushed sample temperatures more closely than the above-ground site (Rack 2), both captured the seasonal pattern well. This, of course, is not necessarily true---in general, a comparison of pipe rack effluent and distributed water temperatures will help to illustrate the limitations of the pipe rack data.

>>>>

@@

Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified. The best sentinel homes, then, are often short-lived. Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with median pH and dissolved inorganic carbon concentrations in pipe rack effluent of 7.3--7.4 and 4--5 mg C L^-1^ (Table 1). But while most water quality parameters were similar across zones, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022]. Aluminum concentrations in Zone 1 were much lower and more consistent throughout the year (Table S1).

@@

Doubling the orthophosphate dose decreased lead release from pipe racks substantially. At the low dose of 1 mg PO~4~ L^-1^, lead in pipe rack effluent did not exhibit a clear time-trend (Figure 1c)---an indication that the pipe racks had stabilized. In particular, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ everywhere (d[Pb]/dt ~ 0, Figure 1a).

@@

Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding Zone 2-specific smooth term (Figure 1a). One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. While this did not impact distributed water pH, it did lower aluminum in treated water (Figure S6). And a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles [@kvech_role_2001]. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999]. But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data we have are not sufficient to explain the trend conclusively.

@@

Temperature in the pipe rack effluent was a reasonable match for distributed water temperature. While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked distributed water and point-of-use flushed sample temperatures more closely than the above-ground site (Rack 2), both captured the seasonal pattern well. This, of course, is not necessarily true---in general, a comparison of pipe rack effluent and distributed water temperatures will help to illustrate the limitations of the pipe rack data.
